PLAE web app enables powerful searching and multiple visualizations across one million unified single-cell ocular transcriptomes

### 2023-02-18 manuscript_00.Rmd
text and figs

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: zotero_plaeApp.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To create a high performance reactive web application to query single cell gene expression data across cell type, species, study, and other factors. METHODS: We updated the content and structure of the underlying data (single cell Eye in a Disk, scEiaD) and wrote the web application PLAE (https://plae.nei.nih.gov) to visualize and explore it. RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes encompassing four species, 60 cell types, six ocular tissues, and 23 body tissues across 37 publications. To demonstrate the value of this unified pan-eye dataset, we replicate known neurogenic and cone macula markers as well as propose six new cone human region markers. CONCLUSION: The PLAE web application provides the eye community a powerful and quick means to test hypotheses related to gene expression across a highly diverse, community-derived database."

@@

  mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
                                      TRUE ~ CellType_predict),
         CellType= case_when(CellType == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
                             TRUE ~ CellType),  
         Tissue = case_when(grepl('Iris', Tissue) ~ 'Iris',
                            TRUE ~ Tissue),
         Tissue = case_when(grepl('^RPE|^Choroid', Tissue) ~ 'RPE-Choroid',
                            TRUE ~ Tissue),
         Citation = case_when(study_accession == 'OGVFB_Hufnagel_iPSC_RPE' ~ 'Swamy VS, Fufa TD, Hufnagel RB, McGaughey DM Building the Mega Single Cell Transcriptome Ocular Meta-Atlas Gigascience 2021 Oct 13;10(10)',
                              TRUE ~ Citation),
         PMID = case_when(study_accession == 'OGVFB_Hufnagel_iPSC_RPE' ~ '34651173',
                          TRUE ~ PMID),
         study_accession = case_when(study_accession == 'OGVFB_Hufnagel_iPSC_RPE' ~ 'SRP329495',
                                     TRUE ~ study_accession))

@@

The outer layer of the eye consists of the cornea and sclera. Beneath the sclera is the uvea which is made of the iris, ciliary body, and the choroid. The inner most layer of the eye is the retina, which is backed by the retinal pigmented epithelium (RPE). The retina alone is composed of six major cell populations: the photoreceptors, horizontal, bipolar, amacrine, ganglion, and non-neuronal cells [@maslandCellPopulationsRetina2011]. These six populations can be further divided into dozens of cell types [@yanCellAtlasHuman2020a]. Furthermore, human disease of the eye can derive from damage to specific cell types. Glaucoma is characterized by damage to the retinal ganglion cells while cone and cone-rod dystrophies are characterized by photoreceptor degeneration. Efforts to untangle the transcriptomics of these diverse cell types with single cell transcriptomics date to the early 2000s when researchers [@hagstromConePigmentGene2000b; @jabsEvidenceP2XP2X2000; @liLocalizationDopamineD1receptor2004; @lindqvistSingleCellRTPCR2002a; @reidRetinoschisinPhotoreceptorsecretedProtein2003]  used RT-PCR approaches to amplify select transcripts in small numbers of cells. Later, researchers used microarray platforms to analyze a broader array of transcripts[@cherryDevelopmentDiversificationRetinal2009; @gustincichGeneDiscoveryGenetically2004a; @kimIdentificationMolecularMarkers2008a; @roeschTranscriptomeRetinalMuller2008; @trimarchiIndividualRetinalProgenitor2008; @trimarchiMolecularHeterogeneityDeveloping2007b; @wahlinMethodAnalysisGene2004a]

The modern era of single cell transcriptomics starts with the introduction of droplet-based technologies, which enabled quantitation of thousands of cells in a single experiment [@kleinDropletBarcodingSingleCell2015]. This approach was pioneered by Macoksko et al. and enabled the profiling of gene expression in 44,808 single cells in the adult mouse retina [@macoskoHighlyParallelGenomewide2015a]. The commercialization of the droplet approach has enabled broad access to this technology and the ocular community have responded by publishing several dozen ocular studies using single cell transcriptome technology since the Macosko publication. An incompmlete list of what the ocular community has used single cell technology across dozens of studies would include studing developmental transcriptomic dynamics, identifying genes to distinguish cell types, proposing finer gradations of cell types within existing categories, comparing and contrasting ocular transcriptomes between organisms, and identify differences in spatially different cells.

Unfortunately manuscripts can only present a sliver of the underlying data and further exploration is laborious, technically challenging, and computationally expensive. Briefly, one would have to identify the relevant publication, find the associated data deposit, download the cell - count matrix, find the cell label table (and optionally the t-SNE or UMAP coordinates), and then load the data into R or python to run the desired queries. If one desired to compare across publications, then one would also have to align the cell type labels and should re-create the count quantification with consistent bioinformatic tooling. It is desirable, perhaps even necessary, to use multiple datasets to assess gene expression to confirm whether a gene expression pattern is consistent and reproducible. Reliance on a single dataset can be problematic as there can be technical issues - for example the large consortium dataset GTEx has persistent contamination from highly expressed genes across tissues [@nieuwenhuisConsistentRNASequencing2020].

Web-based resources to query gene expression can democratize access to large datasets as the computationally intensive steps above can be run once and then shared to anyone with a internet connected device. Single cell RNA-seq web resources that contain ocular tissues include the UCSC Cell Browser, Tabula Sapiens, PanglaoDB, and the Protein Atlas [@consortiumTabulaSapiensSingle2021; @franzenPanglaoDBWebServer2019; @karlssonSingleCellType2021; @speirUCSCCellBrowser2021]. Primary limitations to these resources include incomplete cell type labels and sample metadata that make ocular focused queries challenging. In contrast,the Spectacle resource is an ocular specific web app which has curated over a million cells from dozens of independent resources [@voigtSpectacleInteractiveResource2020]. However each data deposit is independently processed, which makes cross study queries impossible. We recently created a unified ocular data resource which placed dozens of independent single-cell RNA-seq ocular and non-ocular datasets in a single combined space, the Single Cell Eye in a Disk (scEiaD) [@swamyBuildingMegaSinglecell2021].

To broaden the usage of this powerful meta-atlas, we restructured the data into a custom sqlite-based database and built a reactive web application to access it. As many new studies were published since our last data build, we re-ran the scEiaD pipeline in March of 2022. In aggregate we have collated `r meta_filter$study_accession %>% unique() %>% length()` datasets across `r meta_filter$PMID %>% unique() %>% length()` publications, four species, and `r meta_filter$Tissue %>% unique() %>% length()` tissues. We hand-curated `r meta_filter$CellType_predict %>% unique() %>% length()` cell type labels, totaling `r meta_filter %>% filter(!is.na(CellType)) %>% nrow()` cells and used those ground-truth labels as the basis for a machine learning-based algorithm which labels the `r length(meta_filter$CellType_predict[!is.na(meta_filter$CellType_predict)]) - length(meta_filter$CellType[!is.na(meta_filter$CellType)])` more cells. Quality control metrics were tuned and applied to remove lower quality cells. In the end, the single cell Eye in a Disk (scEiaD) database contains `r meta_filter %>% nrow()` cells and 1,509,179,347 non-zero gene/cell expression values. We demonstrate how this unified resource can be used to identify genes that consistently distinguish cell types across studies and species, find genes which have species-specific expression cell type patterning, demonstrate how our pre-computed differential testing can identify markers between similar progenitor cell types, and finally provide a R-based analysis document which end-to-end demonstrates how we use scEiaD to run a custom differential gene expression test to validate and propose new genes that distinguish macula from peripheral cones in humans.

@@

Our full scEiaD pipeline was published previously in Swamy et al. [@swamyBuildingMegaSinglecell2021]. It is composed of three Snakefiles which perform three core workflow tasks[@molderSustainableDataAnalysis2021]. The "SnakeQUANT" pipeline takes the raw fastq input and produces a gene by cell count matrix with kallisto and bustools [@melstedBarcodeUMISet2019; @melstedModularEfficientPreprocessing2019]. This matrix is used in the "SnakePOP" pipeline to identify the optimal parameters for the scANVI based batch correction process (`r supFig_cap(name = 'scPOP_performance', display = 'cite')`). Outputs from this pipeline include the batch corrected lower dimensional space, the UMAP coordinates, cell cluster assignments, and the cell type predictions. The outputs from this pipeline are used in the "SnakeSCEIAD" which performs differential gene testing across multiple conditions and wraps the various data types into the scEiaD sqlite database for use in the plae app (`r supFig_cap(name = 'workflow_network', display = 'cite')`).

@@

To decide whether a droplet is considered a viable cell, we use the "barcodeRanks" algorithm from DropletUtils to automatically identify empty droplets. We then kept cells with the following criteria: less than 20% of the counts were in mitochondrial genes, a cell had more than 600 total counts, at least 300 unique genes quantified, and fewer than 100,000 total counts. To identify likely doublets we used Scrublet [@wolockScrubletComputationalIdentification2019a] and doubletCells (from scran) [@lunStepbystepWorkflowLowlevel2016a] to generate per cell doublet probability scores. To identify cells for removal we used the scores in two ways. First, if an individual cell had a Scrublet score greater than 0.8 (where 1.0 is 100% probability of doublet) and a doubletCell score greater than 1,000,000 (unitless measure where higher value is more likely to be doublet) the cell is labelled as a doublet. Second, if a cluster has an average Scrublet score greater than 0.2 or an average doubletCell score greater than 12 the entire cluster was marked as doublets. These cutoffs were selected by hand inspecting density distributions of the Scrublet and doubletCell scores.

@@

The following alterations were made to Swamy et al.: First, the cutoff to retain a cell was raised from 200 quantified unique genes to 300. Second, we use the DecontX algorithm[@yangDecontaminationAmbientRNA2020] from the celda R package (version 1.9.2) to automatically remove ambient RNA contamination on a per-study basis to produce a updated count matrix. Third, we altered the procedure for aligning gene names across species (see below for details). Fourth, we use the scANVI (scvitools based) batch correction method, which leverages information from the known cell type labels in the batch correction process [@xuProbabilisticHarmonizationAnnotation2021].

A custom cutoff for the minimum required number of unique genes identified per cell were used for the SRP362101 cornea dataset as an abnormally high number of cells were returned from our default value of 300. We set the cutoff for this studies to be 800 as this value resulted in approximately the same number of cells being returned as the authors reported in their papers.

@@

We found in Swamy et al. that scVI had the best batch integration performance relative to other tools like Seurat's CCA, fastMNN, etc. For this study we chose to use a variant of scVI, scANVI, We also found that adjusting the number of latent dimensions, highly variable genes, and epochs had substantial effects on the integration performance. For this update of scEiaD we ran integration with 3000, 4000, 5000, and 6000 highly variable genes, 5 and 15 epochs, and 6, 8, 10, 15, and 20 scANVI outputted latent dimensions. After evaluating the performance of the new scEiaD dataset with our scPOP package we selected the following parameters for the integration with scANVI: 15 latent dimensions, 4000 highly variable genes, and 5 epochs [@swamyBuildingMegaSinglecell2021]. The UMAP was built with a min_dist of 0.1 and the clustering used 50 nearest neighbors.

@@

Every study brought into scEiaD was hand curated to identify unique biological samples, published paper ID (where available), organ, tissue (e.g. Cornea), source (iPSC based or tissue), single cell platform (e.g. 10Xv2, DropSeq, etc.). We also, where possible, annotated retina region (e.g. macula), sex, and age or developmental stage. Samples with perturbations (genetic or treatment) were not included. To import individual cell type labels (e.g. Rod, Muller Glia), we wrote custom code for each study that made this table available to link their cell type assignments back to the matching barcoded cell. We curated two types of cell type labels: "CellType" and "SubCellType" where the former is our normalized cell types taken from the published labels and the later are the original cell type labels. We normalized names for CellType, for example changing "MG" to "Muller Glia" and also dropping more detailed cell type assignments like off and on bipolar cells for Bipolar Cells (as few studies went into this level).

@@

The R package easyPubMed was used to search for 1000 randomly chosen genes present in scEiaD for both gene name alone and gene name plus "AND Retina" with the get_pubmed_ids function. We then used the returned Entrez identifier to pull the abstract with the fetch_pubmed_data function so we could extract the PubMed ID (PMID). This was repeated with the the genes identified as being well supported differentially expressed for ocular cell types. To assess whether the difference in distributions was statistically different, we used the base R t.test function.

@@

Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types [@swamyBuildingMegaSinglecell2021]. We updated the scEiaD database for the PLAE v0.93 web app in six important ways. First, we added a new species, chicken (*Gallus gallus*) [@yamagataCellAtlasChick2021]. Second we added two ocular outflow tract datasets, one brain choroid plexus, and three cornea datasets to enhance coverage across the eye [@collinSingleCellAtlas2021; @gautamMultispeciesSinglecellTranscriptomic2021; @daniCellularSpatialMap2021; @ligockiMolecularCharacteristicsSpatial2021; @patelMolecularTaxonomyHuman2020; @vanzylCellAtlasAqueous2020]. Third, we added a human pan-body reference scRNA dataset to allow for non-ocular comparisons [@heSinglecellTranscriptomeProfiling2020]. Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (`r supFig_cap(name = 'decontX', display = 'cite')`). Fifth, we increased the cell retention cutoff for detected unique transcripts per cell from 200 to 300. Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris project to more easily provide non eye cell type comparisons [@SinglecellTranscriptomics20]. As before, the pipeline's scVI integration parameters were chosen rigorously by testing integration performance across a wide range of latent dimensions and number of highly variable genes in our scPOP framework [@swamyBuildingMegaSinglecell2021] (`r supFig_cap(name = 'scPOP_performance', display = 'cite')`). Each study in the resource was hand assessed to determine whether they had any unusual properties in regards to machine learnt cell type proportions or position in the 2D UMAP space and a small number of outlier studies were removed from the resource.

Of the `r meta_filter %>% filter(Organ == 'Eye') %>% nrow()` ocular cells, only `r meta_filter %>% filter(Organ == 'Eye', !is.na(CellType)) %>% nrow()` had cell type labels we could recover from the publications or data deposits. To label most of the remaining cells we used our xgboost approach from Swamy et al. which we previousy found performed well when compared to other cell type transfer approaches. The overall accuracy is 96% and the confusion matrix shows the predicted cell type to published cell type recall is very high for most cell types (`r supFig_cap(name = 'sup_CT_confusion', display = 'cite')`) and that cell type labelling works well across the different species  (`r supTab_cap(name = 'sup_CT_table', display = 'cite')`). The cell types with lower labelled performance tended to be ciliary, limbal, and corneal cells (which were often labelled as fibroblasts). There are fewer unique datasets for these cell types and we expect that future runs with more data will improve this performance

@@

Several single cell RNA-seq web-based resources contain ocular data (`r tab_cap(name = 'web_resources', display = 'cite')`). The UCSC Cell Browser, Tabula Sapiens portal, PanglaoDB, and Protein Atlas are general purpose single cell repositories that contain some ocular data. The PanglaoDB and UCSC resources have ingressed a large number of published single cell datasets. Af of late 2022, the UCSC Cell Browser has around sixty thousand ocular single cell transcriptomes while the PangloaDB has about forty-seven thousand. However, these resources  do not curate and normalize metadata terms (e.g. ensuring the same naming system for cell types). Furthermore, each study is computationally processed separately. In practice each study is silo-ed off from the rest, making comparisons across studies impractical. The Tabula Sapiens project is an effort to sequence all cell type transcriptomes in humans; while it only contains around around ten thousand ocular cells it can be more readily cross compared to non-ocular tissues / cells. A downside is that the ocular cell type labels can be insufficient - for example rods and cones are together labelled as photoreceptors. The Protein Atlas resource has used only one retina single cell RNA-seq paper to provide cell type label focused information for gene queries (e.g. one can see whether a gene of interest is enriched in Cones). Protein Atlas's use of only one study is problematic as it is unclear whether signal found is specific to the one dataset they use or is common across independent species and datasets.

The Spectacle web resource is the closest comparison to PLAE, as it is a ocular focused resource which has collated a large number of ocular single cell RNA-seq resources (both organoid and tissue based datasets total well over a million cells) and made them available on a web app. PLAE is distinguished from these resources by hand curation of the cell and sample level metadata, re-processing the faw sequence data in a consistent framework, batch correcting the independent studies, and, crucially, providing full availability of the underlying data for outside usage.

@@

web_resources_cap <- tab_cap(name = 'web_resources', caption = 'Comparison of PLAE with other web resources that contain single cell ocular datasets, as of 2021-03-15. The Protein Atlas is asterisked as while it contains independent datasets, it only has one ocular study. Export-able means whether a resource allows bulk downloads of the count matrices and metadata. Organisms refers to the number of ocular species used.')

>>>>

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: zotero_plaeApp.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To create a high performance reactive web application to query single cell gene expression data across cell type, species, study, and other factors. METHODS: We updated the content and structure of the underlying data (single cell Eye in a Disk, scEiaD) and wrote the web application PLAE (https://plae.nei.nih.gov) to visualize and explore it. RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes encompassing four species, 60 cell types, six ocular tissues, and 23 body tissues across 35 publications. To demonstrate the value of this unified pan-eye dataset, we replicate known neurogenic and cone macula markers as well as propose six new cone human region markers. CONCLUSION: The PLAE web application provides the eye community a powerful and quick means to test hypotheses related to gene expression across a highly diverse, community-derived database."

@@

  mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
                                      TRUE ~ CellType_predict),
         CellType= case_when(CellType == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
                             TRUE ~ CellType),  
         Tissue = case_when(grepl('Iris', Tissue) ~ 'Iris',
                            TRUE ~ Tissue),
         Tissue = case_when(grepl('^RPE|^Choroid', Tissue) ~ 'RPE-Choroid',
                            TRUE ~ Tissue),
         Citation = case_when(study_accession == 'OGVFB_Hufnagel_iPSC_RPE' ~ 'Swamy VS, Fufa TD, Hufnagel RB, McGaughey DM Building the Mega Single Cell Transcriptome Ocular Meta-Atlas Gigascience 2021 Oct 13;10(10)',
                              grepl('axon guidance in early-born retinal neurons.', Citation) ~ 'Lo Giudice Q, Leleu M, La Manno G, Fabre PJ. Single-cell transcriptional logic of cell-fate specification and axon guidance in early-born retinal neurons. Development 2019 Sep 9;146(17)',
                              Citation == 'Lu Y, Shiau F, Yi W, Lu S et al. Single-Cell Analysis of Human Retina Identifies Evolutionarily Conserved and Species-Specific Mechanisms Controlling Development. Dev Cell 2020 May 18;53(4):473-491.e9.' ~ 'Lu Y, Shiau F, Yi W, Lu S et al. Single-Cell Analysis of Human Retina Identifies Evolutionarily Conserved and Species-Specific Mechanisms Controlling Development. Dev Cell 2020 May 18;53(4):473-491.e9',
                              TRUE ~ Citation),
         PMID = case_when(study_accession == 'OGVFB_Hufnagel_iPSC_RPE' ~ '34651173',
                          TRUE ~ PMID),
         study_accession = case_when(study_accession == 'OGVFB_Hufnagel_iPSC_RPE' ~ 'SRP329495',
                                     TRUE ~ study_accession))

@@

The outer layer of the eye consists of the cornea and sclera. Beneath the sclera is the uvea which is made of the iris, ciliary body, and the choroid. The inner most layer of the eye is the retina, which is backed by the retinal pigmented epithelium (RPE). Distinct cell types are used to compose each of these layers. The retina alone is composed of six major cell populations: the photoreceptors, horizontal, bipolar, amacrine, ganglion, and non-neuronal cells [@maslandCellPopulationsRetina2011]. These six populations can be subdivided into dozens of cell types [@yanCellAtlasHuman2020a]. Furthermore, human disease of the eye can derive from damage to specific cell types.

Damage to some of these cell types has been implicated in human eye disease. For example glaucoma is characterized by damage to the retinal ganglion cells while cone and cone-rod dystrophies are characterized by photoreceptor degeneration. Efforts to untangle the gene expression patterns of these diverse cell types with single cell transcriptomics date to the early 2000s when researchers [@hagstromConePigmentGene2000b; @jabsEvidenceP2XP2X2000; @liLocalizationDopamineD1receptor2004; @lindqvistSingleCellRTPCR2002a; @reidRetinoschisinPhotoreceptorsecretedProtein2003] used RT-PCR approaches to amplify select transcripts in small numbers of cells. Later, researchers used microarray platforms to analyze a broader array of transcripts[@cherryDevelopmentDiversificationRetinal2009; @gustincichGeneDiscoveryGenetically2004a; @kimIdentificationMolecularMarkers2008a; @roeschTranscriptomeRetinalMuller2008; @trimarchiIndividualRetinalProgenitor2008; @trimarchiMolecularHeterogeneityDeveloping2007b; @wahlinMethodAnalysisGene2004a]

The modern era of single cell transcriptomics started with the introduction of droplet-based technologies, which enabled quantitation of thousands of cells in a single experiment [@kleinDropletBarcodingSingleCell2015]. This approach was pioneered by Macoksko et al. and enabled the profiling of gene expression in 44,808 single cells in the adult mouse retina [@macoskoHighlyParallelGenomewide2015a]. Commercialization of the droplet technology has enabled broad access and the ocular community has responded by using single cell RNA sequencing in several dozen ocular studies.An incomplete list of how the ocular community has used single cell technology would include studying developmental transcriptomic dynamics, identifying genes to distinguish cell types, proposing finer gradations of cell types within existing categories, comparing and contrasting ocular transcriptomes between organisms, and identifying differences in spatially distinct cells.

Unfortunately manuscripts can only present a sliver of the underlying data. Further data exploration can be laborious, technically challenging, and computationally expensive. Briefly, one would have to identify the relevant publication, find the associated data deposit, download the cell by gene count matrix, find the cell label table (and optionally the t-SNE or UMAP coordinates) with the matching fields to align them, and then load the data into R or python to run the desired queries. If one desired to compare across publications, then one would also have to align the cell type label schema and ideally should re-create the count quantification with consistent bioinformatic tooling. It is desirable, perhaps even necessary, to use multiple datasets to assess gene expression to confirm whether a gene expression pattern is consistent and reproducible. Reliance on a single dataset can be problematic as there can be technical issues - for example the large consortium dataset GTEx has persistent contamination from highly expressed genes across tissues [@nieuwenhuisConsistentRNASequencing2020].

Web-based resources to query gene expression can democratize access to large datasets as the computationally intensive steps above can be run once and then shared to anyone with a internet connected device. Single cell RNA-seq web resources that contain ocular tissues include the UCSC Cell Browser, Tabula Sapiens, PanglaoDB, and the Protein Atlas [@consortiumTabulaSapiensSingle2021; @franzenPanglaoDBWebServer2019; @karlssonSingleCellType2021; @speirUCSCCellBrowser2021]. However, ocular focused queries are challenging as these resources have incomplete and inconsistent cell type labels and sample metadata. In contrast to there general purpose resources, Spectacle (http://singlecell-eye.org) is an ocular specific web app which contains over a million cells from dozens of independent resources [@voigtSpectacleInteractiveResource2020]. However each data deposit is independently and inconsistently  processed, which makes cross study queries impossible.

  We recently created a unified ocular data resource which placed dozens of independent single-cell RNA-seq ocular and non-ocular datasets in a single batch corrected space, the Single Cell Eye in a Disk (scEiaD) [@swamyBuildingMegaSinglecell2021]. To broaden the usage of this powerful meta-atlas dataset, we restructured the data into a custom sqlite-based database and built a reactive web application to access it. As many new studies were published since our last data build, we re-ran the scEiaD pipeline in March of 2022. In aggregate we have collated `r meta_filter$study_accession %>% unique() %>% length()` datasets across `r meta_filter$PMID %>% unique() %>% length()` publications, four species, and `r meta_filter$Tissue %>% unique() %>% length()` tissues. We hand-curated `r meta_filter$CellType_predict %>% unique() %>% length()` published cell type labels, totaling `r meta_filter %>% filter(!is.na(CellType)) %>% nrow()` cells and used those ground-truth labels as the basis for a machine learning-based algorithm which labels the `r length(meta_filter$CellType_predict[!is.na(meta_filter$CellType_predict)]) - length(meta_filter$CellType[!is.na(meta_filter$CellType)])` previously unannotated cells. Quality control metrics were tuned and applied to remove lower quality cells. In the end, the single cell Eye in a Disk (scEiaD) database contains `r meta_filter %>% nrow()` cells and 1,509,179,347 non-zero gene/cell expression values. We demonstrate how this unified resource can be used to identify genes that consistently distinguish cell types across studies and species, find genes which have species-specific expression cell type patterning, demonstrate how our pre-computed differential testing can identify markers between similar progenitor cell types, and finally provide a R-based analysis document which end-to-end demonstrates how we use scEiaD to run a custom differential gene expression test to validate and propose new genes that distinguish macula from peripheral cones in humans.

@@

Our full scEiaD pipeline was published previously in Swamy et al [@swamyBuildingMegaSinglecell2021]. It was composed of two Snakefiles that run the workflows [@molderSustainableDataAnalysis2021]. The "SnakeQUANT" workflow takes the raw fastq input and produces a gene by cell count matrix with kallisto and bustools [@melstedBarcodeUMISet2019; @melstedModularEfficientPreprocessing2019]. This matrix is used in the "SnakePOP" pipeline to identify the optimal parameters for the scANVI based batch correction process (`r supFig_cap(name = 'scPOP_performance', display = 'cite')`). Outputs from this pipeline include the batch corrected lower dimensional space, the UMAP coordinates, cell cluster assignments, and the cell type predictions. The outputs from this pipeline are used in a our new Snakefile "SnakeSCEIAD" which performs differential gene testing across multiple conditions and wraps the various data types into the scEiaD sqlite database for use in the plae app (`r supFig_cap(name = 'workflow_network', display = 'cite')`).

@@

To decide whether a droplet is considered a viable cell, we use the "barcodeRanks" algorithm from DropletUtils to automatically identify empty droplets. We then kept cells with the following criteria: less than 20% of the counts were in mitochondrial genes, a cell had more than 600 total counts, at least 300 unique genes quantified, and fewer than 100,000 total counts. To identify likely doublets we used Scrublet [@wolockScrubletComputationalIdentification2019a] and doubletCells (from scran) [@lunStepbystepWorkflowLowlevel2016a] to generate per cell doublet probability scores. To identify cells for removal we used the scores in two ways. First, if an individual cell had a Scrublet score greater than 0.8 (where 1.0 is 100% probability of doublet) and a doubletCell score greater than 1,000,000 (unitless measure where higher value is more likely to be doublet) the cell is labelled as a doublet. Second, if a cluster has an average Scrublet score greater than 0.2 or an average doubletCell score greater than 12 the entire cluster was marked as doublets. All of these cutoff values were selected by hand inspecting density distributions of the Scrublet and doubletCell scores.

@@

The following alterations were made to the Swamy et al. pipeline: First, the cutoff to retain a cell was raised from 200 quantified unique genes to 300. Second, we use the DecontX algorithm[@yangDecontaminationAmbientRNA2020] from the celda R package (version 1.9.2) to automatically remove ambient RNA contamination on a per-study basis to produce a updated count matrix. Third, we altered the procedure for aligning gene names across species (see below for details). Fourth, we use the scANVI (scvitools based) batch correction method, which leverages information from the known cell type labels in the batch correction process [@xuProbabilisticHarmonizationAnnotation2021].

A custom cutoff for the minimum required number of unique genes identified per cell were used for the SRP362101 cornea dataset as an abnormally high number of cells were returned from our default value of 300. We set the cutoff for this study to 800 as this value resulted in approximately the same number of cells being returned as the authors reported in their paper.

@@

We found in Swamy et al. that scVI had the best batch integration performance relative to other tools like Seurat's CCA, fastMNN, harmony, ComBat, scanorama, and others. We previously found that adjusting the number of latent dimensions, highly variable genes, and epochs had substantial effects on the integration performance. For this update of scEiaD we ran integration with 3000, 4000, 5000, and 6000 highly variable genes, 5 and 15 epochs, and 6, 8, 10, 15, and 20 scANVI outputted latent dimensions. After evaluating the performance of the new scEiaD dataset with our scPOP package we selected the following parameters for the integration: 15 latent dimensions, 4000 highly variable genes, and 5 epochs [@swamyBuildingMegaSinglecell2021]. The UMAP was built with a min_dist of 0.1 and the clustering used 50 nearest neighbors.

@@

Every study brought into scEiaD was hand curated to identify the unique biological samples, published paper ID (where available), organ, tissue (e.g. Cornea), source (iPSC based or tissue), single cell platform (e.g. 10Xv2, DropSeq, etc.). We also, where possible, annotated retina region (e.g. macula), sex, and age or developmental stage. Samples with perturbations (genetic or treatment) were not included. To import individual cell type labels (e.g. Rod, Muller Glia), we wrote custom code for each study that made this table available to link their cell type assignments back to the matching barcoded cell. We curated two types of cell type labels: "CellType" and "SubCellType" where the former is our normalized cell types taken from the published labels and the later are the original cell type labels. We normalized names for CellType, for example changing "MG" to "Muller Glia" and also dropping more detailed cell type assignments like off and on bipolar cells for Bipolar Cells (as few studies went into this level).

@@

The R package easyPubMed was used to search for 1000 randomly chosen genes present in scEiaD for both gene name alone and gene name plus "AND Retina" with the get_pubmed_ids function . We then used the returned Entrez identifier to pull the abstract with the fetch_pubmed_data function so we could extract the PubMed ID (PMID). This was repeated with the the genes identified as being well supported differentially expressed for ocular cell types. To assess whether the difference in distributions was statistically different, we used the base R t.test function.

@@

Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` datasets, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types [@swamyBuildingMegaSinglecell2021]. We updated the scEiaD database for the PLAE v0.93 web app in six important ways. First, we added a new species, chicken (*Gallus gallus*) [@yamagataCellAtlasChick2021]. Second we added two ocular outflow tract datasets, one brain choroid plexus, and three cornea datasets to enhance coverage across the eye [@collinSingleCellAtlas2021; @gautamMultispeciesSinglecellTranscriptomic2021; @daniCellularSpatialMap2021; @ligockiMolecularCharacteristicsSpatial2021; @patelMolecularTaxonomyHuman2020; @vanzylCellAtlasAqueous2020]. Third, we added a human pan-body reference scRNA dataset to allow for non-ocular comparisons [@heSinglecellTranscriptomeProfiling2020]. Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (`r supFig_cap(name = 'decontX', display = 'cite')`). Fifth, we increased the stringency of the cell retention cutoff for detected unique transcripts per cell from 200 to 300. Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris project to more easily provide non eye cell type comparisons [@SinglecellTranscriptomics20]. As before, the pipeline's scVI integration parameters were chosen rigorously by testing integration performance across a wide range of latent dimensions and number of highly variable genes in our scPOP framework [@swamyBuildingMegaSinglecell2021] (`r supFig_cap(name = 'scPOP_performance', display = 'cite')`).

Of the `r meta_filter %>% filter(Organ == 'Eye') %>% nrow()` ocular cells, only `r meta_filter %>% filter(Organ == 'Eye', !is.na(CellType)) %>% nrow()` had cell type labels we could recover from the publications or data deposits. To label most of the remaining cells we used our xgboost approach from Swamy et al. which we previousy found performed well when compared to other cell type transfer approaches. The overall accuracy is 96% and the confusion matrix shows the predicted cell type to published cell type recall is very high for most cell types (`r supFig_cap(name = 'sup_CT_confusion', display = 'cite')`). We also see that the cell type labelling works well across the different species  (`r supTab_cap(name = 'sup_CT_table', display = 'cite')`). The cell types with lower labelling performance tended to be ciliary, limbal, and corneal cells (which were often labelled as fibroblasts). There are fewer unique datasets for these cell types and we expect that future runs with more data will improve this performance

@@

Several single cell RNA-seq web-based resources contain ocular data (`r tab_cap(name = 'web_resources', display = 'cite')`). The UCSC Cell Browser, Tabula Sapiens portal, PanglaoDB, and Protein Atlas are general purpose single cell repositories that contain some ocular data. The PanglaoDB and UCSC resources have ingressed a large number of published single cell datasets. As of late 2022, the UCSC Cell Browser had around sixty thousand ocular single cell transcriptomes while the PangloaDB has about forty-seven thousand. However, these resources  do not curate and normalize metadata terms. This means that cell types have inconsistent naming which makes cross study comparisons very challenging. The Tabula Sapiens project is an effort to sequence all cell type transcriptomes in humans; while it only contains around around ten thousand ocular cells it can be more readily cross compared to non-ocular tissues / cells. A downside is that the ocular cell type labels can be insufficient - for example rods and cones are together labelled as photoreceptors. The Protein Atlas resource has used only one retina single cell RNA-seq paper to provide cell type label focused information for gene queries (e.g. one can see whether a gene of interest is enriched in Cones). Protein Atlas's use of only one study is problematic as it is unclear whether signal found is specific to the one dataset they use or is common across independent species and datasets.

The Spectacle web resource is the closest comparison to PLAE, as it is a ocular focused resource which has collated a large number of ocular single cell RNA-seq resources (both organoid and tissue based datasets total well over a million cells) and made them available on a web app. PLAE is distinguished from these resources by hand curation of the cell and sample level metadata, re-processing the raw sequence data in a consistent framework, batch correcting the independent studies, and, crucially, providing full availability of the underlying data for outside usage.

@@

web_resources_cap <- tab_cap(name = 'web_resources', caption = 'Comparison of PLAE with other web resources that contain single cell ocular datasets, as of 2021-03-15. The Protein Atlas is asterisked as while it contains independent datasets, it only has one ocular study. Export-able means whether a resource allows bulk downloads of the count matrices and metadata. Organisms refers to the number of ocular species with ocular data')
